---
title: "Properties of Exponential Distribution Means"
author: "Craig Struble"
date: "November 9, 2014"
output: pdf_document
---

# Introduction
The exponential distribution
is a probability distribution
generally used
to model the time between
events governed
by a Poisson process.
The exponential distribution
has a single parameter
called `rate`
and represented
by the symbol $\lambda$,
and its probability density function
$f(x)$ is defined as
$$f(x) = \lambda e^{-\lambda x}.$$

In this report, the properties of sample means
from exponential distributions are investigated
via simulation using the R statistical programming
language and system.
The report begins with the code
for creating the simulated data
in the `Simulation` section.
Visualizations and analysis
of the simulations
are presented
in the `Results and Discussion`
section.

# Simulation
The instructions
for this report stated
that exponential
with $\lambda = 0.2$
and sample size
from the exponential
of $40$.
The number
of simulated samples
from our exponential
distribution
should be at least $1000$,
but $10000$ was chosen
for a better illustration.
Variables are
initialized with those values.

```{r setup, echo=FALSE}
# Setup report
library(ggplot2)
library(knitr)
set.seed(141109)
```

```{r params}
# Parameters for the report
lambda <- 0.2
sampleSize <- 40
numDist <- 10000
```

The simulations are executed
the requested number of times.
The sample mean
from each simulation is
stored
in a variable 
for plotting later.

```{r samplemeans}
# Generate sample means for samples drawn from an exponential distribution
mns <- sapply(1:numDist, function(x) mean(rexp(sampleSize, lambda)))
```

# Results and Discussion
In the previous section,
the sample means
from $`r numDist`$
samples of size $`r sampleSize`$
from the exponential distribution
with rate $`r lambda`$
were generated.
In this section,
three properties
of the sample mean distribution
are presented and compared
with theoretical expectations:
the center of the distribution;
the variance of the distribution;
the normality of the distribution.

## Center of Sample Mean Distribution
The distribution
of sample means should be
centered around $1/\lambda = 5$,
the population mean of the exponential distribution.

The mean of the sample means
is $`r mean(mns)`$,
which is $`r abs(mean(mns) - 1/lambda)`$
from the theoretical expectation.
The following figure plots a histogram
of the generated sample means,
with a vertical line at the theoretical expectation
of $`r 1/lambda`$.

```{r samplemeanhist,echo=FALSE}

# Plot histogram of sample means
df <- data.frame(means=mns)
g <- ggplot(df, aes(x=means)) + 
    geom_histogram(binwidth=0.1, color="darkgreen", fill="white")
g + geom_vline(xintercept=1/lambda, size=3, alpha=0.5) +
    labs(x="Sample Means", title="Histogram of Exponential Sample Means")
```

## Variance of Sample Mean Distribution
The expected variance of the sample mean distribution
is $\sigma^2 / n$, where $\sigma^2$ is the variance
of the exponential distribution and $n$ is the sample
size of the simulated samples.
The variance of an exponential distribution
is $1 / \lambda^2$, which is $`r 1/lambda^2`$
for a rate of `r lambda`.

Filling in the actual values
in our simulations,
the expected variance is $`r (1/lambda^2) / sampleSize`$.
The observed variance of the sample mean distribution is $`r var(mns)`$.
The absolute difference
in observed versus expected variance
is $`r abs(var(mns) - ((1/lambda^2) / sampleSize))`$. 

## The Sample Mean Distribution is Approximately Normal
According to theoretical expectations,
the distribution
of sample means
from any distribution
should be
well approximated by a normal distribution.
The theoretical normal distribution
for the sample mean distribution
is
$$\bar{X}_{n} \sim N(\mu, \sigma^2/n).$$
Substituting in for our simulations,
the sample means should be approximated
by
$$\bar{X}_{40} \sim N(5, 25/40).$$

To begin this section,
a figure is presented
to show that the distribution
of $`r numDist`$ exponentially distributed
values does not match
our expectation of a normal
distribution, but
that its sample mean
does fall within the theoretical normal.
The red curve shows the theoretical normal
density curve
and the gray line shows the sample mean
for the exponentially distributed
random values.

```{r expnotnormal, echo=FALSE, fig.height=4}

# Plot random values from an exponential distribution
df <- data.frame(val=rexp(numDist, lambda))
g <- ggplot(df, aes(x=val)) + 
    geom_histogram(aes(y=..density..), binwidth=0.5, color="darkgreen", fill="white")
g + labs(x="Value from Exponential", 
         title="Histogram of Exponential Random Values") +
  stat_function(fun=dnorm, 
                args=list(mean=1/lambda, sd=1/(lambda*sqrt(sampleSize))), 
                color="red", size=3, alpha=0.5) +
    geom_vline(xintercept=mean(df$val), size=3, alpha=0.5)
```

The next figure shows that the distribution 
of sample means
closely follows the theoretical normal distribution.
The theoretical normal distribution
is shown again as a red curve.
The mean of sample means,
shown as a gray line,
is near the peak
of the theoretical normal distribution.

```{r samplemeansnormal, echo=FALSE, fig.height=4}

# Plot density of sample means and compare to theoretical normal
df <- data.frame(means=mns)
g <- ggplot(df, aes(x=means)) + 
    geom_histogram(aes(y=..density..), binwidth=0.1, color="darkgreen", 
                   fill="white")
g + geom_vline(xintercept=1/lambda, size=3, alpha=0.5) +
    labs(x="Sample Means", title="Histogram of Exponential Sample Means") +
    stat_function(fun=dnorm, 
                  args=list(mean=1/lambda, sd=1/(lambda*sqrt(sampleSize))), 
                  color="red", size=3, alpha=0.5) +
    geom_vline(xintercept=mean(mns), size=3, alpha=0.5)
```

# Appendix: Full Code
This appendix contains the full code listing for this report.

```{r, ref.label=all_labels(), echo=TRUE, eval=FALSE}
```