---
title: "An Analysis of the ToothGrowth Data Set in R"
author: "Craig Struble"
date: "November 14, 2014"
output:
    pdf_document:
        number_sections: true
        fig_caption: true
---

# Introduction
Data
on the effect
of vitamin C
on tooth growth
is available
in R
as the ToothGrowth data set.
The study
looked
at three different
doses
of vitamin C
with two delivery methods
and measured
the length
of teeth
in 10 guinea pigs.
In this report,
the delivery methods
and doses are compared
to determine
if vitamin C
has an impact
on tooth growth.

```{r startup}
library(ggplot2)
library(knitr)
```

# Exploratory Analysis and Summary

The `ToothGrowth` data set
is part of the `datasets` package.
The package 
and data
are loaded.

```{r setup}
library(datasets)
data(ToothGrowth)
```

The data consists
of `r nrow(ToothGrowth)` rows
and `r ncol(ToothGrowth)` columns.
The `len` column contains the tooth lengths.
The `supp` column contains the delivery method
or supplement type,
which is represented as a factor
with levels `OJ` and `VC`
representing orange juice
and ascorbic acid respectively.
The `dose` column is numeric
containing the three dose levels: 0.5, 1, and 2 mg.

```{r summary}
str(ToothGrowth)
summary(ToothGrowth)
```

The data is plotted comparing
tooth length by dose,
grouped by supplement type.
A linear fit
to the data is provided
in the plot.

```{r exploratoryplot, echo=FALSE}
g <- ggplot(ToothGrowth, aes(x=dose, y=len, colour=supp))
g + geom_point() + geom_smooth(method=lm) +
    labs(x="Dose (mg)", y="Tooth Length (units unknown)", title="ToothGrowth Data Set") +
    guides(col=guide_legend(title="Supplement"))
```

# Comparison of Delivery Method and Dosage

# Conclusions

# Appendix: Full Code
This appendix contains the full code listing for this report.

```{r, ref.label=all_labels(), echo=TRUE, eval=FALSE}
```