---
title: "More Vitamin C Increases Guinea Pig Tooth Growth: An Analysis in R"
author: "Craig Struble"
date: "November 14, 2014"
output:
    pdf_document:
        number_sections: true
        fig_caption: true
---

# Introduction
In this report,
delivery methods
and doses are compared
to determine
if vitamin C
has an impact
on tooth growth
using the `ToothGrowth`
data in R.
The original study
looked
at three doses
with two delivery methods.
Tooth length
was measured
in 60 guinea pigs,
10 at each combination 
of treatments.
The analysis herein shows
that
vitamin C dose amounts have a significant impact
on tooth growth.
Delivery methods have a significant impact
at lower dose levels, but not at higher doses.

The report has three additional sections.
Section 2 explores and summarizes the data.
Section 3 performs hypothesis testing on variables
to identify significant influences on tooth growth.
Section 4 draws conclusions from the testing performed in the report.

```{r startup, echo=FALSE}
# Intial setup, load necessary packages.
library(ggplot2)
library(knitr)

```

# Exploratory Analysis and Summary

The `ToothGrowth` data set
is part of the `datasets` package.
The package 
and data
are loaded.

```{r setup}
# Load the data
library(datasets)
data(ToothGrowth)
```

The data consist
of `r nrow(ToothGrowth)` rows
and `r ncol(ToothGrowth)` columns.
The `len` column contains the tooth lengths.
The `supp` column contains the delivery method
or supplement type,
which is represented as a factor
with levels `OJ` and `VC`
representing orange juice
and ascorbic acid respectively.
The `dose` column is numeric
containing the three dose levels: 0.5, 1, and 2 mg.
A plot comparing
tooth length by dose,
grouped by supplement type,
follows the summary tables.

```{r summary}
# Summarize the data
str(ToothGrowth)
summary(ToothGrowth)
```

```{r exploratoryplot, echo=FALSE, fig.width=6, fig.height=3, fig.cap="The tooth length plotted against dosage amount. Different supplement types are plotted by color. Linear fits with confidence intervals are plotted for each group. A strong association between dosage amount and tooth length appears to exist for both supplement types."}

# Plot the tooth length by dose, grouped by supplement
g <- ggplot(ToothGrowth, aes(x=dose, y=len, colour=supp))
g + geom_point() + geom_smooth(method=lm) +
    labs(x="Dose (mg)", y="Tooth Length (units unknown)", title="ToothGrowth Data Set") +
    guides(col=guide_legend(title="Supplement"))
```

# Comparison of Delivery Method and Dosage

The `ToothGrowth` data set has two variables
of interest: the supplement type `supp`,
and the dosage amount `dose`.
In Figure 1,
two relationships appear to exist.
The first is that increased dose amounts
correspond to increased tooth length.
The second is that orange juice
may be a better supplement
for delivering vitamin C
than ascorbic acid.
Each is investigated
using statistical methods
to determine if there is a likely
relationship.

An important aspect
of the data
is that there are 60 distinct
guinea pigs, 10 at each combination
of variables 
(see original data source
Crampton, The Journal of Nutrition, vol. 33, issue 5, May 1947, pp. 491-504).
Hence, paired analyses
are not appropriate
and are not used.

The p-value
threshold
for significance is assumed
to be less than 0.05.

## Comparison of Dosages

To compare dosages,
T tests are executed
for each pair of dosage
levels.
It is unknown whether
vitamin C dose affects
tooth length variance,
so unequal variances are assumed.

```{r ttestdoseallpairs}
# All pair T tests for doses
t.test(len ~ dose, data=subset(ToothGrowth, dose %in% c(0.5, 1.0)))
t.test(len ~ dose, data=subset(ToothGrowth, dose %in% c(0.5, 2.0)))
t.test(len ~ dose, data=subset(ToothGrowth, dose %in% c(1.0, 2.0)))
```

In each case,
the null hypothesis
that the means are equal
is rejected at $p < 0.05$.
Since $3$ tests were considered, 
a Bonferroni correction is made
for multiple testing.
The corrected
level of significance
is $0.05/3 = 0.017$.
All tests remain significant
with $p < 0.017$,
suggesting that dose amounts
change tooth length.

## Comparison of Delivery Methods

To compare delivery methods,
initially a T test
of all the data is performed.
It is unknown whether
vitamin C delivery method affects
tooth length variance,
so unequal variances are assumed.

```{r ttestalldelivery}
# T test for delivery, ignoring doses. Assigned to variable for inline reference
tst <- t.test(len ~ supp, data=ToothGrowth)
tst
```

The p-value of `r tst$p.value` does not reach the level
of significance of 0.05 
to reject the null hypothesis that the means are equal.

Figure 1 does suggest some difference
in delivery methods at each dose level.
T tests are performed
for the delivery methods
at each dose level individually.

```{r ttestdeliverydose}
# T tests for delivery methods, at each dosage level.
t.test(len ~ supp, data=subset(ToothGrowth, dose == 0.5))
t.test(len ~ supp, data=subset(ToothGrowth, dose == 1.0))
t.test(len ~ supp, data=subset(ToothGrowth, dose == 2.0))
```

A total
of 4 tests were performed
to evaluate significance,
so a Bonferroni correction
of $0.05/4=0.0125$
is used as the level
of significance needed
to reject the null hypothesis.

For the doses 0.5 mg and 1.0 mg,
the null hypothesis of equal means
is rejected,
suggesting that the vitamin C supplement type
is an important factor in tooth length.
For 2.0 mg,
the null hypothesis cannot be rejected,
and indeed the means appear to be close 
in the groups.

# Conclusions
In the analysis,
the following assumptions were made:
the samples are not paired;
the samples have unequal variance
since test conditions could change variance;
and
the threshold
for significance is 0.05 before multiple test corrections.



Using T tests,
tooth length
increases
with increased
vitamin C doses
($p < 0.017$, Bonferroni corrected).

T tests
were also used
to compare vitamin C delivery methods.
Orange juice appears
to be a more effective delivery method
than ascorbic acid
for lower doses (0.5 mg and 1.0 mg, $p < 0.0125$, Bonferroni corrected).
No significant difference between delivery methods was seen
at the highest dose, 2.0 mg.

# Appendix: Full Code
This appendix contains the full code listing for this report.

```{r, ref.label=all_labels(), echo=TRUE, eval=FALSE}
```